Unshuffling Permutations

### 2015-09-17 article.tex
Diverses petites modif

%%
%%
%% ---- Introduction ----
%%
\section{Introduction}
\label{section:Introduction}
The {\em shuffle product} $\shuffle$ is a very famous operation on words
first defined by Eilenberg and MacLane~\cite{Eilenberg:MacLane:1953}.
Given three words $u$, $v$, and $w$, $u$ is said to be a \emph{shuffle}
of $v$ and $w$ if it can be formed by interleaving the letters from $v$
and $w$ in a way that maintains the left-to-right ordering of the
letters from each word. In addition to purely combinatorial questions,
the shuffle product of words leads to the following computational ones:
\begin{enumerate}
    \item \label{item:problem_1}
    Given two words $v_1$ and $v_2$, compute the set $v_1 \shuffle v_2$.
    \item \label{item:problem_2}
    Given three words $u$, $v_1$, and $v_2$, decide if $u$ is a shuffle
    of $v_1$ and $v_2$.
    \item \label{item:problem_3}
    Given words $u$, $v_1$, \dots, $v_k$, decide if $u$ is in
    $v_1 \shuffle \dots \shuffle v_k$.
    \item \label{item:problem_4}
    Given a word $u$, decide if there exist two words $v_1$ and $v_2$
    such that $u$ is a shuffle of $v_1$ and $v_2$.
\end{enumerate}
Even if these problems seem similar, they are very different in terms
of complexity. Let us now review some facts about these. In what follows,
$n$ denotes the size of $u$ and $m_i$ denotes the size of each $v_i$.
A solution to Problem~\ref{item:problem_1} can be computed in
\begin{equation}
    O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)
\end{equation}
time~\cite{Spehner:TCS:1986}. An improvement and a generalization
of~Problem~\ref{item:problem_1} has been proposed
in~\cite{Allauzen:IGM:2000}, where it is proved that, given words
$v_1$, \dots, $v_k$, the iterated shuffle
$v_1 \shuffle \dots \shuffle v_k$ can be computed in
\begin{equation}
    O\left(\binom{m_1 + \dots + m_k}{m_1, \dots, m_k}\right)
\end{equation}
time. Problem~\ref{item:problem_2} is in \Pclass; it is indeed a
classical textbook exercise to design an efficient dynamic programming
algorithm solving it. It can be tested in $O\left(n^2 / \log(n)\right)$
time~\cite{Leeuwen:Nivat:IPL:1982}. To the best of our knowledge, the
first $O(n^2)$ time algorithm for this problem appeared
in~\cite{Mansfield:DAM:1983}. This algorithm can easily be extended to
check in polynomial-time whether or not a word is in the shuffle of any
fixed number of given words. Nevertheless, Problem~\ref{item:problem_3}
is \NPC~\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}. This
remains true even if the ground alphabet has size
$3$~\cite{Warmuth:Haussler:JCSS:1984}. Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that
Problem~\ref{item:problem_3} remains \NPC even if all the words $v_i$,
$i \in [k]$, are identical, thereby proving that, for two words $u$ and
$v$, it is \NPC to decide whether or not $u$ is in the iterated shuffle
of $v$. Again, this remains true even if the ground alphabet has size $3$.
Let us now focus on Problem~\ref{item:problem_4}. It is shown
in~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013} that it is \NPC to
decide if a word $u$ is a \emph{square} (w.r.t. the shuffle), that is
a word $u$ with the property that there exists a word $v$ such that $u$
is a shuffle of $v$ with itself. The hardness of this last particular
problem implies that Problem~\ref{item:problem_4} is \NPC.
\smallskip

This paper is intended to study a natural generalization of $\shuffle$,
denoted by $\SHUFFLE$, as a shuffle of permutations. Roughly speaking,
given three permutations $\pi$, $\mu$, and $\nu$, $\pi$ is said to be a
{\em shuffle} of $\mu$ and $\nu$ if $\pi$ (viewed as a word) is a
shuffle of two words that are order-isomorphic to $\mu$ and $\nu$.
Our intention in this paper is to study this shuffle product of
permutations $\SHUFFLE$ both from a combinatorial and from a
computational point of view by focusing on {\em square} permutations,
that are permutations $\pi$ being in the shuffle of a permutation $\mu$
with itself. The shuffle product defined and considered in this paper
is different from other existing and well-known shuffle products on
permutations. For instance, in~\cite{DHT:IJAC:2002}, the authors define
the {\em shifted shuffle product}. For this product, $\pi$ is a shuffle
of $\mu$ and $\nu$ if $\pi$ is in the shuffle, as words, of $\mu$ and
the word obtained by incrementing all the letters of $\nu$ by the size
of $\mu$. It is a simple exercise to prove that, given three
permutations $\pi$, $\mu$, and $\nu$, deciding if $\pi$ is a in the
shifted shuffle of $\mu$ and $\nu$ is in~\Pclass.
\smallskip

This paper is organized as follows. In
Section~\ref{section:Shuffle product on permutations} we provide a
precise definition of $\SHUFFLE$. This definition passes through the
preliminary definition of an operator $\Delta$ to unshuffle permutations.
This operator is in fact a coproduct, endowing the space of all
permutations with a coalgebra structure. By duality, the unshuffling
operator $\Delta$ leads to the definition of our shuffle operation on
permutations. This approach has many advantages. First, some
combinatorial properties of $\SHUFFLE$ depend on properties of $\Delta$
and are more easy to prove on the coproduct side. Second, this way of
doing allows to obtain a clear description of the multiplicities of the
elements appearing in the shuffle of two permutations, which are worthy
of interest. Section~\ref{section:Binary square words and permutations}
is devoted to show that the problems related to the shuffle of words has
links with the shuffle of permutations. In particular, we show that
binary words that are square are in one-to-one bijection with square
permutations
(Proposition~\ref{prop:bijection_binary_to_permutations_squares}). Next,
Section~\ref{section:Algebraic issues} presents some algebraic and
combinatorial properties of~$\SHUFFLE$. In particular, we show that
$\SHUFFLE$ is associative and commutative
(Proposition~\ref{prop:shuffle_associative_commutative}), and that
if a permutation if a square, its mirror, complement, and inverse also
are square (Proposition~\ref{prop:square_stability}). Finally,
Section~\ref{section:Algorithmic issues} presents
the most import result of this paper: the fact that deciding if a
permutation is a square is \NPC (Proposition~\ref{proposition:hardness}).

@@

\begin{proposition} \label{prop:endomorphisms}
    The three linear maps
    \begin{equation}
        \phi_1, \phi_2, \phi_3 : \QQ[S] \to \QQ[S]
    \end{equation}
    linearly sending a permutation $\pi$ to, respectively,
    $\widetilde{\pi}$, $\bar \pi$, and $\pi^{-1}$ are endomorphims of
    associative algebras.
\end{proposition}
\begin{proof}[of Proposition~\ref{prop:endomorphisms}]
    To prove, for $j = 1, 2, 3$, that $\phi_j$ is a morphism of
    associative algebras, we have to prove that for all permutations
    $\sigma$ and $\nu$,
    \begin{equation}
        \phi_j(\sigma \SHUFFLE \nu) =
        \phi_j(\sigma) \SHUFFLE \phi_j(\nu).
    \end{equation}
    We shall proceed here as we have done during the proof of
    Proposition~\ref{prop:shuffle_associative_commutative} by showing
    that $\phi_j$ is a morphism of coalgebras, that is
    \begin{equation}
        \Delta \phi_j = (\phi_j \otimes \phi_j) \Delta.
    \end{equation}
    In the sequel, $\pi$ is a permutation.
    \smallskip

@@

We now use the algebraic properties of $\SHUFFLE$ exhibited by
Proposition~\ref{prop:endomorphisms} to obtain combinatorial properties
of square permutations.
\medskip

\begin{proposition} \label{prop:square_stability}
    Let $\pi$ be a square permutation and $\rho$ be a square root of
    $\pi$. Then,
    \begin{enumerate}[label={\it (\roman*)},fullwidth]
        \item \label{item:square_stability_1}
        the permutation $\widetilde{\pi}$ is a square and
        $\widetilde{\rho}$ is one of its square roots;
        \item \label{item:square_stability_2}
        the permutation $\bar \pi$ is a square and $\bar \rho$ is one of
        its square roots;
        \item \label{item:square_stability_3}
        the permutation $\pi^{-1}$ is a square and $\rho^{-1}$ is one of
        its square roots.
    \end{enumerate}
\end{proposition}
\begin{proof}[of Proposition~\ref{prop:square_stability}]
    All statements~\ref{item:square_stability_1},
    \ref{item:square_stability_2}, and~\ref{item:square_stability_3} are
    consequences of Proposition~\ref{prop:endomorphisms}. Indeed,
    since $\pi$ is a square permutation and $\rho$ is a square root of
    $\pi$, by definition, $\pi$ appears in the product
    $\rho \SHUFFLE \rho$. Now, by Proposition~\ref{prop:endomorphisms},
    for any $j = 1, 2, 3$, since $\phi_j$ is a morphism of associative
    algebras from $\QQ[S]$ to $\QQ[S]$, $\phi_j$ commutes with the
    shuffle product of permutations $\SHUFFLE$. Hence, in particular,
    one has
    \begin{equation}
        \phi_j(\rho \SHUFFLE \rho) =
        \phi_j(\rho) \SHUFFLE \phi_j(\rho).
    \end{equation}
    Then, since $\pi$ appears in $\rho \SHUFFLE \rho$, $\phi_j(\pi)$
    appears in $\phi_j(\rho \SHUFFLE \rho)$ and appears also in
    $\phi_j(\rho) \SHUFFLE \phi_j(\rho)$. This shows that
    $\phi_j(\rho)$ is a square root of $\phi_j(\pi)$ and
    implies~\ref{item:square_stability_1}, \ref{item:square_stability_2},
    and~\ref{item:square_stability_3}.
    \qed
\end{proof}
\medskip

>>>>

%%
%%
%% ---- Introduction ----
%%
\section{Introduction}
\label{section:Introduction}
The {\em shuffle product} $\shuffle$ is a very famous operation on words
first defined by Eilenberg and Mac Lane~\cite{Eilenberg:MacLane:1953}.
Given three words $u$, $v_1$, and $v_2$, $u$ is said to be a \emph{shuffle}
of $v_1$ and $v_2$ if it can be formed by interleaving the letters from
$v_1$ and $v_2$ in a way that maintains the left-to-right ordering of the
letters from each word. In addition to purely combinatorial questions,
the shuffle product of words leads to the following computational ones:
\begin{enumerate}
    \item \label{item:problem_1}
    Given two words $v_1$ and $v_2$, compute the set $v_1 \shuffle v_2$.
    \item \label{item:problem_2}
    Given three words $u$, $v_1$, and $v_2$, decide if $u$ is a shuffle
    of $v_1$ and $v_2$.
    \item \label{item:problem_3}
    Given words $u$, $v_1$, \dots, $v_k$, decide if $u$ is in
    $v_1 \shuffle \dots \shuffle v_k$.
    \item \label{item:problem_4}
    Given a word $u$, decide if there exist two words $v_1$ and $v_2$
    such that $u$ is a shuffle of $v_1$ and $v_2$.
\end{enumerate}
Even if these problems seem similar, they are very different in terms
of complexity. Let us now review some facts about these. In what follows,
$n$ denotes the size of $u$ and $m_i$ denotes the size of each $v_i$.
A solution to Problem~\ref{item:problem_1} can be computed in
\begin{equation}
    O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)
\end{equation}
time~\cite{Spehner:TCS:1986}. An improvement and a generalization
of~Problem~\ref{item:problem_1} has been proposed
in~\cite{Allauzen:IGM:2000}, where it is proved that, given words
$v_1$, \dots, $v_k$, the iterated shuffle
$v_1 \shuffle \dots \shuffle v_k$ can be computed in
\begin{equation}
    O\left(\binom{m_1 + \dots + m_k}{m_1, \dots, m_k}\right)
\end{equation}
time. Problem~\ref{item:problem_2} is in \Pclass; it is indeed a
classical textbook exercise to design an efficient dynamic programming
algorithm solving it. It can be tested in $O\left(n^2 / \log(n)\right)$
time~\cite{Leeuwen:Nivat:IPL:1982}. To the best of our knowledge, the
first $O(n^2)$ time algorithm for this problem appeared
in~\cite{Mansfield:DAM:1983}. This algorithm can easily be extended to
check in polynomial-time whether or not a word is in the shuffle of any
fixed number of given words. Nevertheless, Problem~\ref{item:problem_3}
is \NPC~\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}. This
remains true even if the ground alphabet has size
$3$~\cite{Warmuth:Haussler:JCSS:1984}. Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that
Problem~\ref{item:problem_3} remains \NPC even if all the words $v_i$,
$i \in [k]$, are identical, thereby proving that, for two words $u$ and
$v$, it is \NPC to decide whether or not $u$ is in the iterated shuffle
of $v$. Again, this remains true even if the ground alphabet has size $3$.
Let us now focus on Problem~\ref{item:problem_4}. It is shown
in~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013} that it is \NPC to
decide if a word $u$ is a \emph{square} (w.r.t. the shuffle), that is
a word $u$ with the property that there exists a word $v$ such that $u$
is a shuffle of $v$ with itself. The hardness of this last particular
problem implies that Problem~\ref{item:problem_4} is \NPC.
\smallskip

This paper is intended to study a natural generalization of $\shuffle$,
denoted by $\SHUFFLE$, as a shuffle of permutations. Roughly speaking,
given three permutations $\pi$, $\sigma_1$, and $\sigma_2$, $\pi$ is
said to be a {\em shuffle} of $\sigma_1$ and $\sigma_2$ if $\pi$ (viewed
as a word) is a shuffle of two words that are order-isomorphic to
$\sigma_1$ and $\sigma_2$. Our intention in this paper is to study this
shuffle product of permutations $\SHUFFLE$ both from a combinatorial and
from a computational point of view by focusing on {\em square}
permutations, that are permutations $\pi$ being in the shuffle of a
permutation $\sigma$ with itself. The shuffle product defined and
considered in this paper is different from other existing and well-known
shuffle products on permutations. For instance, in~\cite{DHT:IJAC:2002},
the authors define the {\em shifted shuffle product}. For this product,
$\pi$ is a shuffle of $\sigma_1$ and $\sigma_2$ if $\pi$ is in the
shuffle, as words, of $\sigma_1$ and the word obtained by incrementing
all the letters of $\sigma_2$ by the size of $\sigma_1$. It is a simple
exercise to prove that, given three permutations $\pi$, $\sigma_1$, and
$\sigma_2$, deciding if $\pi$ is a in the shifted shuffle of $\sigma_1$
and $\sigma_2$ is in~\Pclass.
\smallskip

This paper is organized as follows. In
Section~\ref{section:Shuffle product on permutations} we provide a
precise definition of $\SHUFFLE$. This definition passes through the
preliminary definition of an operator~$\Delta$, allowing to
{\em unshuffle} permutations. This operator is in fact a coproduct,
endowing the space of all permutations with a coalgebra structure
(see~\cite{Joni:Rota:1979} or~\cite{Grinberg:Reiner:2014} for the
definition of these algebraic structures). By duality, the unshuffling
operator $\Delta$ leads to the definition of our shuffle operation on
permutations. This approach has many advantages. First, some
combinatorial properties of $\SHUFFLE$ depend on properties of $\Delta$
and are more easy to prove on the coproduct side. Second, this way of
doing allows to obtain a clear description of the multiplicities of the
elements appearing in the shuffle of two permutations, which are worthy
of interest from a combinatorial point of view.
Section~\ref{section:Binary square words and permutations} is devoted to
show that the problems related to the shuffle of words has links with
the shuffle of permutations. In particular, we show that binary words
that are square are in one-to-one bijection with square permutations
avoiding some patterns
(Proposition~\ref{prop:bijection_binary_to_permutations_squares}). Next,
Section~\ref{section:Algebraic issues} presents some algebraic and
combinatorial properties of~$\SHUFFLE$. We show that $\SHUFFLE$ is
associative and commutative
(Proposition~\ref{prop:shuffle_associative_commutative}), and that if a
permutation is a square, its mirror, complement, and inverse are also
squares (Proposition~\ref{prop:square_stability}). Finally,
Section~\ref{section:Algorithmic issues} presents the most import result
of this paper: the fact that deciding if a permutation is a square is
\NPC (Proposition~\ref{proposition:hardness}).

@@

\begin{lemma} \label{lem:endomorphisms}
    The three linear maps
    \begin{equation}
        \phi_1, \phi_2, \phi_3 : \QQ[S] \to \QQ[S]
    \end{equation}
    linearly sending a permutation $\pi$ to, respectively,
    $\widetilde{\pi}$, $\bar \pi$, and $\pi^{-1}$ are endomorphims of
    associative algebras.
\end{lemma}
\begin{proof}[of Lemma~\ref{lem:endomorphisms}]
    To prove, for $j = 1, 2, 3$, that $\phi_j$ is a morphism of
    associative algebras, we have to prove that for all permutations
    $\sigma$ and $\nu$,
    \begin{equation}
        \phi_j(\sigma \SHUFFLE \nu) =
        \phi_j(\sigma) \SHUFFLE \phi_j(\nu).
    \end{equation}
    We shall proceed here as we have done during the proof of
    Proposition~\ref{prop:shuffle_associative_commutative} by showing
    that $\phi_j$ is a morphism of coalgebras, that is
    \begin{equation}
        \Delta \phi_j = (\phi_j \otimes \phi_j) \Delta.
    \end{equation}
    In the sequel, $\pi$ is a permutation.
    \smallskip

@@

We now use the algebraic properties of $\SHUFFLE$ exhibited by
Lemma~\ref{lem:endomorphisms} to obtain combinatorial properties
of square permutations.
\medskip

\begin{proposition} \label{prop:square_stability}
    Let $\pi$ be a square permutation and $\rho$ be a square root of
    $\pi$. Then,
    \begin{enumerate}[label={\it (\roman*)},fullwidth]
        \item \label{item:square_stability_1}
        the permutation $\widetilde{\pi}$ is a square and
        $\widetilde{\rho}$ is one of its square roots;
        \item \label{item:square_stability_2}
        the permutation $\bar \pi$ is a square and $\bar \rho$ is one of
        its square roots;
        \item \label{item:square_stability_3}
        the permutation $\pi^{-1}$ is a square and $\rho^{-1}$ is one of
        its square roots.
    \end{enumerate}
\end{proposition}
\begin{proof}[of Proposition~\ref{prop:square_stability}]
    All statements~\ref{item:square_stability_1},
    \ref{item:square_stability_2}, and~\ref{item:square_stability_3} are
    consequences of Lemma~\ref{lem:endomorphisms}. Indeed,
    since $\pi$ is a square permutation and $\rho$ is a square root of
    $\pi$, by definition, $\pi$ appears in the product
    $\rho \SHUFFLE \rho$. Now, by Lemma~\ref{lem:endomorphisms},
    for any $j = 1, 2, 3$, since $\phi_j$ is a morphism of associative
    algebras from $\QQ[S]$ to $\QQ[S]$, $\phi_j$ commutes with the
    shuffle product of permutations $\SHUFFLE$. Hence, in particular,
    one has
    \begin{equation}
        \phi_j(\rho \SHUFFLE \rho) =
        \phi_j(\rho) \SHUFFLE \phi_j(\rho).
    \end{equation}
    Then, since $\pi$ appears in $\rho \SHUFFLE \rho$, $\phi_j(\pi)$
    appears in $\phi_j(\rho \SHUFFLE \rho)$ and appears also in
    $\phi_j(\rho) \SHUFFLE \phi_j(\rho)$. This shows that
    $\phi_j(\rho)$ is a square root of $\phi_j(\pi)$ and
    implies~\ref{item:square_stability_1}, \ref{item:square_stability_2},
    and~\ref{item:square_stability_3}.
    \qed
\end{proof}
\medskip
